---
title: "Getting Started"
author: ""
date: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Welcome to this guide on how to work with `ctsmTMB` for modelling time-series data.

In this example we consider a simple 1D mean-reverting stochastic differential equation model, the Ornstein-Uhlenbeck process:
$$ 
\mathrm{d}x_{t} = \theta (\mu - x_{t}) \, \mathrm{d}t \, + \sigma_{x} \, \mathrm{d}b_{t} 
$$
Here $\theta$, $\mu$ and $\sigma_x$ are (fixed effects) parameters to be estimated. 
  
We assume that we have **direct** observations of the state $x_{t}$, at discrete times $t_k$ for $k=0,1,2,...,N$ i.e. the observation equation is
$$
y_{t_{k}} = x_{t_{k}} + \varepsilon_{t_{k}}
$$

The residuals are assumed to be Gaussian, and we choose the variance as
$$
\varepsilon_{t_{k}} \sim N(0,\sigma_{y}^{2} \cdot u_{t_{k}})
$$


Here $\sigma_{y}$ is a fixed effect parameter (also to be estimated), and $u_{t}$ is a known time-dependent input. The input is added here for sake of introduction, just to demonstrate how time-dependent inputs are specified. We could for example imagine the input to be given by
$$
u_{t_{k}} = \left\{ 1, 4, 1, 4, 1, 4 \cdots \right\}
$$
such that certain observations have twice the standard deviation of others, and thus carry less weight (information) about the latent state $x$. In the remainder of this vignette however, we simply set $u_{t} = 1$.

# Initialising
  
We initialise a `ctsmTMB` model object using

```{r}
library(ctsmTMB)
model <- ctsmTMB$new()
```

Printing the model object in the console reveals some basic information about it:
```{r}
print(model)
```

# Add system equations

---

First we specify the stochastic differential equation governing our latent state $x_t$. This is straight-forward using **R** formulas, choosing appropriate character names for the parameters.
```{r}
model$addSystem(dx ~ theta * (mu - x) * dt + sigma_x * dw)
```

We emphasize that drift terms must be multiplied by a `dt` and diffusion terms by a `dw` or `dw#` where # is some number e.g. `dw2`. A single state equation may contain any number of diffusion terms i.e.

```{r, eval=FALSE}
model$addSystem(dx ~ theta * (mu - x) * dt + sigma_x1 * dw1 + x * sigma_x2 * dw2 + dw3)
```

# Add observation equations

---

Now the observation equations of the system must be specified. This amounts to specifying the mean and variance of the residual normal distribution. First the mean is specified i.e.
```{r}
model$addObs(y ~ x)
```

The variable used on the left-hand side, here `y`, determines the name of the relevant observations to-be provided in the data for maximum likelihood estimation later.

# Add observation variances

---

Next we specify the residual variance, using the name given to the observation equation above, in `addObs`:
```{r}
model$setVariance(y ~ sigma_y^2 * u)
```

# Add inputs

---

Next, we declare which variables are time-dependent inputs via
```{r}
model$addInput(u)
```

These shall must also be provided in the data later, similar to observations.


# Add parameters

---

We must also specify the (fixed effects) parameters, and in addition their initial/lower/upper values for the estimation.
```{r}
model$setParameter(
  theta   = c(initial = 5,    lower = 0,    upper = 20),
  mu      = c(initial = 0,    lower = -10,  upper = 10),
  sigma_x = c(initial = 1e-1, lower = 1e-5, upper = 5),
  sigma_y = c(initial = 1e-1, lower = 1e-5, upper = 5)
)
```

A parameter can be fixed by supplying just a single value. It is for instance typically useful to fix observation noise parameters because simultaneous identification of observation and process noise is difficult in practice, and some knowledge about observation noise may be known. Thus, let us fix $\sigma_{y}$ to a somewhat arbitrary but reasonable value:
```{r}
model$setParameter(
  sigma_y  = 0.05
)
```

Let's inspect the model object again, and see that it is no longer empty:
```{r}
print(model)
```

# Set initial state and covariance

---

Lastly the state distribution at the initial time point must be specified via its mean and variance.
```{r}
initial.state <- list(mean=1, cov=1e-1)
model$setInitialState(initial.state=initial.state)
```

Note that in higher dimensions the provided covariance `cov` must be a matrix. A simple initial covariance could just be a scaled identity e.g. in two dimensions `cov = 1e-1 * diag(2)`

# Generate Data

---

The model has now been fully specified, and state and parameter estimation can be performed on the data at hand.

In this particular example we generate some fake data. This can be achieved by simulating a stochastic realization of the stochastic differential equation, and adding some observation noise to it.

We acheive this task using the  `simulate` method of the model object, which perform stochatic simulations based on the [Euler-Maruyama scheme](https://en.wikipedia.org/wiki/Euler–Maruyama_method). The user is referred to the [simulation vignette](https://phillipbvetter.github.io/ctsmTMB/articles/simulate.html) for further details.

We choose the true parameters to be
$$
\theta = 10.0 \qquad \mu=1.00 \qquad \sigma_{x} = 1.00 \qquad \sigma_{y}=0.05
$$

The code below performs the simulation and prepares data for likelihood estimation:
```{r, fig.height=5,fig.width=9, out.width="100%", fig.align='center'}
library(ggplot2)

# Set simulation settings
set.seed(11)
true.pars <- c(theta=10, mu=1, sigma_x=1, sigma_y=0.05)
dt.sim <- 1e-3
t.end <- 1
t.sim <- seq(0, t.end, by=dt.sim)
df.sim <- data.frame(t=t.sim, u=1, y=NA)

# Perform simulation
sim <- model$simulate(data=df.sim, 
                      pars=true.pars, 
                      n.sims=1, 
                      silent=T, 
                      initial.state=initial.state)
x <- sim$states$x$i0$x1

# Extract observations from simulation and add noise
iobs <- seq(1,length(t.sim), by=10)
t.obs <- t.sim[iobs]
y = x[iobs] + true.pars["sigma_y"] * rnorm(length(iobs))

# Create data-frame
data <- data.frame(
  t = t.obs,
  u = 1,
  y = y
)

# Plot the simulation and observed data
ggplot() +
  geom_line(aes(x=t.sim,y=x,color="Simulation")) +
  geom_point(aes(x=t.obs,y=y,fill="Observations")) +
  ctsmTMB:::getggplot2theme() + labs(x="t", y="x",color="",fill="")
```

# Perform estimation

---

We can now pass the `data` to the `estimate` method. This will build the model, perform various checks, construct the computational graph for automatic differentiation, and then perform the optimization.

_**Note**: The data *must* contain an increasing time column named `t` and columns for each of the specified inputs and observations, in this case `u` and `y`._

```{r}
fit <- model$estimate(data)
```

_**Note**: a time-consuming step in the estimation procedure is construction of the AD graph of the underlying likelihood function, but the time spent for this task relative to the optimization time will even out for models with more parameters._

The output generated during the optimization is the objective (negativ log-likelihood) value and the parameter values at the current step. The optimizer used by **ctsmTMB** is the *nlminb* optimizer from the *stats* library.

We refer the user to the [estimation vignette](https://phillipbvetter.github.io/ctsmTMB/articles/estimate.html) for further details on the available arguments to `estimate`, and to the ['use another optimizer' vignette](https://phillipbvetter.github.io/ctsmTMB/articles/using_another_optimizer.html) for details on how to use another optimizer than *nlminb*.

# Inspecting the `fit` object

---

The `fit` object contains the following entries
```{r}
names(fit)
```

The first is a boolean which indicates whether estimation was succesful
```{r}
fit$convergence
```
which is just a copy of the optimization message from `stats::nlminb`.

The second is the likelihood value at the found optimum
```{r}
fit$nll
```

the third is the likelihood gradient at the optimum
```{r}
fit$nll.gradient
```

and the fourth is the likelihood hessian at the optimum
```{r}
fit$nll.hessian
```

## Parameter estimates

---

Printing the `fit` object reveals a standard coefficient matrix for the parameter estimates.

```{r}
print(fit)
```

We can see the parameter estimate and the associated standard error together with a one-dimensional t-test statistic and associated P-value for the common hypothesis test
$$
H_{0}: p = 0 \\
H_{1}: p \neq 0
$$

_**Note** The large uncertainty in $\theta$ here is primarily caused by a relatively short time-series (2 seconds), relative to the characteristic time of the process $\tau = 1/\theta = 0.1 \, \text{sec}$._

The parameter-related information can be extracted from the fit object. The estimated (fixed) parameters:
```{r}
fit$par.fixed
```

The standard deviations of the (fixed) parameters:
```{r}
fit$sd.fixed
```

The covariance of the (fixed) parameters:
```{r}
fit$cov.fixed
```

The parameter covariance is found by inverting the likelihood hessian at the found optimum i.e.:
```{r}
solve(fit$nll.hessian)
```

## State estimates

---

The optimal state distributions associated with the estimated parameters can be extracted from the model object as well.

In this example we used the Extended Kalman filter (this is the default filtering algorithm specified by the argument `method="ekf"` to `estimate`). This method produces  *prior* and *posterior* state estimates. The prior estimates are one-step-ahead predictions, while the posterior estimated are these priors but updated against the current-time available observations. The user is referred to the [Kalman Filter vignette](https://phillipbvetter.github.io/ctsmTMB/articles/extended_kalman_filter.html) for more information on the theoretical details.

```{r, fig.height=5,fig.width=9, out.width="100%", fig.align='center', eval=F,include=F}
# Extract time, and prior/posterior mean and standard deviation estimates
t         = fit$states$mean$posterior$t
xprior    = fit$states$mean$prior$x
xprior_sd = fit$states$sd$prior$x
xpost     = fit$states$mean$posterior$x
xpost_sd  = fit$states$sd$posterior$x

# Create vector c(xprior[1], xpost[1], xprior[2], xpost[2],...)
t2 <- c(rbind(t,t))
xprior_post <- c(rbind(xprior, xpost))
xprior_post_sd <- c(rbind(xprior_sd,xpost_sd))

# Plot
ggplot() +
  geom_line(aes(x=t2,y=xprior_post,color="State Estimates (Prior-Posterior)"),lwd=1) +
  geom_point(aes(x=data$t,data$y, color="Observations")) +
  labs(x = "Time", y = "", color="") +
  ctsmTMB:::getggplot2theme()
```
```{r eval=FALSE, include=FALSE}
  # geom_line(aes(x=t,y=xpost,color="State Estimates (Posterior)"),lwd=1) +
  # geom_line(aes(x=t,y=xprior,color="State Estimates (Prior)"),lwd=1) +
  # geom_ribbon(aes(x=t,ymin=xpost-2*xpost_sd,ymax=xpost+2*xpost_sd),fill="grey",alpha=0.5) +
```

The states can easily be plotted using the provided **S3** plot.ctsmTMB.fit method. Here we plot both prior and posterior states against the observations:

```{r, fig.height=5,fig.width=7, out.width="100%", fig.align='center'}
plot(fit, type="states",state.type="prior",against="y")
```
```{r, fig.height=5,fig.width=7, out.width="100%", fig.align='center'}
plot(fit, type="states",state.type="posterior",against="y")
```

_**Note:** The decrease in variance for the posterior state estimate is expected because these states are updated to the observations._

<!-- We can quickly perform a comparison with the *smoothed* state estimates from the `laplace` estimation method (we suppress the output): -->
<!-- ```{r, fig.height=5,fig.width=7, out.width="100%", fig.align='center'} -->
<!-- fit2 <- model$estimate(data, method="laplace", silent=TRUE, control=list(trace=0)) -->
<!-- plot(fit2, type="states", state.type="smoothed", against="y") -->
<!-- ``` -->


## Residual analysis

---

Model validation typically involves inspecting the properties of prediction residuals. The model residuals are contained in `fit$residuals` with the entries:

```{r}
names(fit$residuals)
```

We can also easily generate a reisdual analysis plot with the **S3** plot.ctsmTMB.fit method. This is actually the default arguments to `plot.ctsmTMB.fit`. This produces a time-series of the residuals, a histogram, a quantile-quantile plot, auto/partial-correlations and a cumulative periodogram:

```{r, fig.height=9,fig.width=9, out.width="100%", fig.align='center'}
plot(fit)
```

# Profiling the likelihood

---

We can perform likelihood profiling with the `profile` S3 method on the `fit` object, and further plot the result by calling the `plot` S3 method on that.

For an example we can inspect the profile likelihood of $\theta$ as follows:
```{r, fig.height=5,fig.width=9, out.width="100%", fig.align='center'}
a <- fit$par.fixed["theta"] - 5
b <- fit$par.fixed["theta"] + 5
prof <- profile(fit, list("theta"=seq(a,b,length.out=25)), silent=TRUE)
plot(prof)
```

# **Extra:** Algebraic equations

---

The model definitions can be kept clean by defining algebraic expressions which replace variables in the defined equations. A typical scenario where algebraic equations can be used is to rename parameters which must be strictly positive. 

**Example** In this model $\theta$ should be a strictly positive parameter $\hat{\theta} = \exp(\log\theta)$. This can be achieved by setting the following algrabic expression:
```{r}
model$setAlgebraics(theta ~ exp(logtheta))
```

The effect of this is to replace all occurences of `theta` in the model equations with `exp(logtheta)`. The final thing to do is to add a new parameter entry to the model object which describes values for `logtheta` 
```{r}
model$setParameter(logtheta = log(c(initial=5, lower=0, upper=20)))
```

