---
title: "Moment Predictions"
author: ""
date: ""
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Moment Predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
\newcommand{\E}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\V}[1]{\mathrm{V}\left[#1\right]}
\newcommand{\diff}{\mathrm{d}}
\newcommand{\dp}[2]{\dfrac{\partial #1}{\partial #2}}
\newcommand{\dd}[2]{\dfrac{\diff #1}{\diff #2}}
  
<style>
body {
  text-align: justify;
  hyphens: auto;
}
</style>
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, cache=FALSE)
```

```{r set-options, echo=FALSE}
options(width=1e2)
```

```{r, eval=FALSE, include=FALSE}
# We can use this to type 'rmd' in the console, and the markdown html document is then knitted fast.
render.rmd("rmd")
```

This vignette demonstrates how to use the `predict` method for calculating *k-step-ahead* forecasts for the mean and variance, available for the Kalman filtering methods (`lkf`, `ekf` and `ukf`).

**Note:** These moment predictions assume that the underlying probability density of the SDE remains approximately Gaussian. This will hold for linear and mildly non-linear SDEs. In addition the assumption will be worse the longer the prediction horizon. In those cases the `simulate` method may be more useful (see the relevant [vignette](https://phillipbvetter.github.io/ctsmTMB/articles/simulate.html)).

## Introduction

### Notation
- We use subscript $x_{t} = x(t)$ to denote time for continuous variables.

- We use subscript $y_k = y(t_k)$ to indicate discrete time-points for discrete variables.

- We denote the set of observations from the initial time $t_0$ until the "current" time $t_{i}$ by
$$
\mathcal{Y}_{i} = \left\{ y_{i}, y_{i-1},...,y_{1},y_{0} \right\}
$$

- We denote mean and variance by $\mu_{t} = \E{x_t}$ and $P_{t} = \V{x_t}$.

- We use $\mu_{k \mid k-1}$ and $P_{k \mid k-1}$ to denote the mean/variance at time $t_{k}$ conditional on the observations $\mathcal{Y}_{k-1}$ i.e.
$$
\mu_{k \mid k} = \E{x_{t_k} \mid \mathcal{Y}_{k}} \qquad\qquad P_{k \mid k} = \V{x_{t_k} \mid \mathcal{Y}_{k}}
$$


### Stochastic State Space System

We consider the following types of stochastic state space systems

$$
x_{t} = f(t,x_t,u_t,\theta) \, \mathrm{d}t + G(t,x_t,u_t, \theta) \, \diff B_{t}
$$

$$
y_{k} = h(t_k, x_{t_k}, u_{t_k}, \theta) + \varepsilon_{k}
$$

where the observation noise is zero-mean Gaussian $\varepsilon_{k} \sim \mathcal{N}(0,\Sigma(t_k, x_{t_k}, u_{t_k}, \theta))$. In this notation $u$ are inputs and $\theta$ are fixed effects parameters to be estimated. 

We refer to the functions $f$ as the *drift*, $G$ as the *diffusion*, and $h$ as the *link*. We may omit arguments and just write e.g. $f(x_t)$ for readability.

### Observations and Likelihood

The likelihood, which is the joint density of all observations, can be rewritten by using repeated conditioning as

$$
L(\theta) = p(\mathcal{Y}_{N}) = \prod_{k=1}^{N} p(y_{k} \mid \mathcal{Y}_{k-1}) = \prod_{k=1}^{N} p(h(x_{k}) + \varepsilon_{k} \mid \mathcal{Y}_{k-1}) 
$$

### What is a "Prediction"?

When we say *predictions* we mean forecasts of the mean and variance (first and second order moments) of the underlying SDE. The fact that these are used to represent the forecast distribution implies the underlying assumption of linearity of the SDE, such that distributions are always Gaussian.

We use the moment differential equations of the SDE to generate these forecasts.

### Moment Differential Equations

The moment differential equations (MDEs) are ordinary differential equations that govern the evolution of the moments of $x_{t}$. The first two moments (mean and variance) are governed by

$$
\dd{\mu}{t} = \E{f(x_t)}
$$

$$
\dd{P}{t}  = \E{ f(x_{t})\left(x_t - \E{x_t}\right)^{T} } + \E{\left(x_t - \E{x_t}\right)f(x_{t})^{T} } + \E{ G(x_t) G(x_t)^{T} }
$$

In the approximate Gaussian setting where, with respect to $x$, $f$ is linear and $G$ is constant, the MDEs simplify to

$$
\dd{\mu}{t} = f(\mu_t)
$$

$$
\dd{P}{t} = \left[\dp{f}{x}(\mu_t)\right] P + P \left[\dp{f}{x}(\mu_t)\right]^{T} + G(\mu_t) \left[G(\mu_t)\right]^{T}
$$

A *k-step* prediction is a prior estimate of the assumed Gaussian distribution *k* time-steps ahead, relatively to the current position. This distribution is parametrized by the prior mean $\mu_{i+k \mid i}$ and variance $P_{i+k \mid i}$. These priors are obtained by integrating the MDEs forward in time, starting from the current posterior distribution parametrized by $\mu_{i \mid i}$ and $P_{i \mid i}$ i.e.

$$
\mu_{i+k|i} = \mu_{i|i} + \int_{t_{i}}^{t_{i+k}} f(\mu_{t}) \, dt
$$

$$
P_{i+k|i} = P_{i|i} + \int_{t_{i}}^{t_{i+k}} 
\left(
\left[\dp{f}{x}(\mu_t)\right] P + 
P \left[\dp{f}{x}(\mu_t)\right]^{T} + 
G(\mu_t) \left[G(\mu_t)\right]^{T} 
\right)
\, dt
$$

In a Kalman filter setting this amounts to repeatedly performing time-update steps while skipping the data-update step entirely.

### Algorithm

When using the `predict` method the forecast horizon is controlled by
the `k.ahead` argument, i.e. how many time-steps "into the future"
forecasts are wanted for. The algorithm returns $N$ (`N = nrow(data) - k.ahead`) forecast scenarios. Each of these $i=0,1,..,N-1$ scenarios consist of
$k_{\text{ahead}}+1$ state values, one for each time point $t=t_{i+j}$ where
$j=0,1,..,k_{\text{ahead}}$.

The algorithm carries out the following step-wise procedure:

-   Filter with the provided data to obtain posterior state and variance estimates for every point in time in the provided `data[,"t"]` column.

-   Extract the first $N$ estimates and discard the remainder.

-   For each $i=0,1,..,N-1$ posteriors, integrate the MDEs forward from the initial time $t_{i}$ until time $t_{i+k_{\text{ahead}}}$ producing moment forecasts at each of the intermediate times $t_{i+j}$.

The output of `predict` is a list with the two entries `states` and `observations`. Each entry is a matrix with $N \cdot (1+k_{\text{ahead}})$ rows, obtained by row-binding each forecast scenario together.



## Example

We consider the Ornstein Uhlenbeck SDE with time-varying mean and with identify link function:

$$ 
\diff x_{t} = \theta (a_t - x_{t}) \, \diff t \, + \sigma_{x} \, \diff b_{t}
$$

$$
y_{t_{k}} = x_{t_{k}} + \varepsilon_{t_{k}}
$$

The mean is the "arbitrary" time-varying signal $a_t = tu_{t}^{2}-\cos(tu_{t})$, where $u_{t}$ is a known input signal.

### Create Model

First we create the model as follows:

```{r}
## Load libraries
library(ctsmTMB)
library(ggplot2) ## plots
library(latex2exp) ## LaTeX plot fonts

## Create model
model <- newModel()
model$addSystem(dx ~ theta * (a*u^2-cos(2*pi*t*u) - x) * dt + sigma_x*dw)
model$addObs(y ~ x)
model$setVariance(y ~ sigma_y^2)
model$addInput(u)

## Set parameter values
## note: not strictly necessary to set lower/upper bounds
model$setParameter(
  theta   = c(initial = 2, lower = 0,    upper = 100),
  a       = c(initial = 1, lower = 1e-5, upper = 100),
  sigma_x = c(initial = 0.2, lower = 1e-5, upper = 5),
  ## fix sigma_y to 0.05 by not giving any upper/lower bounds
  sigma_y = c(initial = 5e-2)
)

## Set initial state mean and covariance
## note: diag(1) is not strictly needed
model$setInitialState(list(1, 1e-1*diag(1)))
```

### Create Data using Simulate

Now we create data to perform predictions on using the `simulate` method, using the true parameters

$$
\theta = 20 \qquad a = 1 \qquad \sigma_{x} = 1.00 \qquad \sigma_{y}=0.05
$$
and the input signal

$$
u_{t} = \tanh\left(\sin\left(B_{t}\right)\right)
$$
```{r}
## set true parameters, and create data
true.pars <- c(theta=4, a=10, sigma_x=1, sigma_y=0.05)
dt.sim <- 1e-3
t.sim <- seq(0, 1, by=dt.sim)

## seed for input creation
set.seed(20)
u.sim <- cumsum(rnorm(length(t.sim),sd=0.1))
# u.sim <- tanh(sin(u.sim))
df.sim <- data.frame(t=t.sim, y=NA, u=u.sim)

## set rng seeds for C++ states and observations
cpp.seeds <- c(20,20)

## perform simulation
sim <- model$simulate(data=df.sim, 
                      pars=true.pars, 
                      n.sims=1,
                      silent=T,
                      cpp.seeds = cpp.seeds)

## extract simulations
y.sim <- sim$observations$y$i0[,1]

## extract observations as every 10th simulation
iobs <- seq(1, length(t.sim), by=10)
t.obs <- t.sim[iobs]
y.obs <- y.sim[iobs]
u.obs <- u.sim[iobs]

## create data
df.obs <- data.frame(
  t = t.obs,
  u = u.obs,
  y = y.obs
)
```

We show the simulated and observed inputs and observations in the plot below.

```{r, echo=TRUE, fig.height=5,fig.width=9,out.width="100%", fig.align='center'}
ggplot() + 
  geom_line(aes(x=t.sim, y=y.sim, color="y_t")) +
  geom_point(aes(x=t.obs, y=y.obs, color="y_k")) +
  geom_line(aes(x=t.sim, y=u.sim, color="u_t")) +
  geom_point(aes(x=t.obs, y=u.obs, color="u_k")) +
  scale_color_discrete(
    labels=unname(TeX(c("$y_{t}$","$y_{k}$","$u_{t}$","$u_{k}$"))),
    breaks = c("y_t", "y_k", "u_t", "u_k"),
  ) +
  ctsmTMB:::getggplot2theme() +
  theme(legend.text = ggplot2::element_text(size=15)) +
  labs(color="",x="Time",y="")
```

### Predict

We can now try to predict on the generated data

```{r}
## set the first observation to NA
df.temp <- df.obs
df.temp$y[1] <- NA

## predict
pred <- model$predict(data=df.temp,
                      pars=true.pars,
                      k.ahead=nrow(df.temp)-1)
```


A few notes for the code above:

-   We use the default value of argument `k.ahead = nrow(data)-1` to get
a forecast over the entire time-vector. This means that there is
only $N=1$ forecast scenario, but its very long.

-   We remove the first observation to prevent a posterior update at the
initial time-point `t[1]`. This allows us to fully control the
initial state mean and variance via `setInitialState`.

-   Since `k.ahead = nrow(data)-1` and $N=1$ implies just one forecast
scenario starting from the initial time point the remaining
observations `y[2:nrow(data)]` are not used at all, and might as well be `NA`.

### Output

The returned `pred` object is a list of two matrices named `states` and `observations`.

1. The first five columns of each matrix are time indices and time-values for each of the forecast scenarios:
    - The column `i` holds the time-index from where the forecast began.
    - The column `j` holds the time-index for where the forecast is.
    - The column `t.i` is the numeric time value associated with index `i.`
    - The column `t.j` is the numeric time value associated with index `j.`
    - The column `k.ahead` is the number of forecasted timesteps since the last posterior update (if any data was available at that time-point).

2. The next columns in both matrices are mean and covariances for the states and observations respectively, named appropriately.

3. The final columns in the `observations` matrix also hold the corresponding observations in the data. The name is suffixed by `.data` e.g. `y.data`.

```{r}
head(pred$states)

head(pred$observations)
```


### Long prediction horizons

We can perform long horizon estimates simply by setting the `k.ahead` argument very large. The default and maximum allowed value is `k.ahead = nrow(.data)-1`. Thus, by default, the `predict` method will generate one single prediction the length of the provided data.

This is useful for instance when searching for good initial guesses for the optimizer before calling the `estimate` method. Consider the example below where we calculate predictions for a series of differnet values of `theta`:

```{r, message=FALSE}
## create vectors with different theta's
pars <- lapply(c(1,2,4,8), function(x) c(x, 10, 1, 0.05))

## predict for the various theta values
pred1 = model$predict(df.obs, pars=pars[[1]])
pred2 = model$predict(df.obs, pars=pars[[2]])
pred3 = model$predict(df.obs, pars=pars[[3]])
pred4 = model$predict(df.obs, pars=pars[[4]])
```

Plotting these state estimates against the data we get a feeling that $\theta \in \left[10,50\right]$ (with $\theta=20$ being the truth here).

```{r, echo=TRUE, fig.height=5,fig.width=9,out.width="100%", fig.align='center'}
plot.df <- data.frame(t = pred$states[,"t.j"],
                      t.obs = t.obs,
                      y.obs = y.obs,
                      x1=pred1$states[,"x"],
                      x2=pred2$states[,"x"],
                      x3=pred3$states[,"x"],
                      x4=pred4$states[,"x"])
ggplot(data=plot.df) +
  geom_line(aes(x=t, y=x1, color="x1")) +
  geom_line(aes(x=t, y=x2, color="x2")) +
  geom_line(aes(x=t, y=x3, color="x3")) +
  geom_line(aes(x=t, y=x4, color="x4")) +
  geom_point(aes(x=t.obs, y=y.obs, color="y")) +
  scale_color_discrete(
    labels=unname(TeX(c("$\\theta=1$","$\\theta=10$","$\\theta=50","$\\theta=100$","$y_{k}$"))),
    breaks = c("x1", "x2", "x3", "x4", "y"),
  ) +
  ctsmTMB:::getggplot2theme() + 
  theme(legend.text = ggplot2::element_text(size=15)) +
  labs(color="",x="Time",y="")
```

### Forecast Evaluation

We can evaluate the forecast performance of our model by comparing predictions against the observed data. We start by estimating the most likely parameters of the model:

```{r}
fit = model$estimate(df.obs)

print(fit)
```


and then predict over an appropriate forecast horizon. In this example we let that horizon be 25-steps:
```{r}
pred.horizon <- 15
pred = model$predict(df.obs, k.ahead=pred.horizon)
```

Let's plot the `pred.horizon`-step predictions against the observations.
```{r}
pred.obs = pred$states[pred$states[,"k.ahead"]==pred.horizon,]
```


```{r, echo=FALSE, fig.height=5,fig.width=9,out.width="100%", fig.align='center', fig.alt='The plot shows 25-step predictions with associated 95% prediction interval in grey, against the observations.'}
t.j <- pred.obs[,"t.j"]
x <- pred.obs[,"x"]
x.sd <- sqrt(pred.obs[,"var.x"])

ggplot() +
  geom_line(aes(x=t.j, y=x,color="25-Step Predictions")) +
  geom_ribbon(aes(x=t.j, ymin=x-2*x.sd, ymax=x + 2*x.sd),fill="grey",alpha=0.5) +
  geom_point(aes(x=t.obs, y.obs,color="Observations")) +
  labs(color="",x="Time",y="") +
  ctsmTMB:::getggplot2theme()
```

Lastly we calculate the mean prediction accuracy using an RMSE-score for all prediction horizons:
```{r}
pdf <- data.frame(pred$states, y.data = pred$observations[,"y.data"])
rmse <- data.frame(
  k.ahead = 0:pred.horizon,
  rmse = sapply(
    with(pdf, split(pdf, k.ahead)), 
    function(df) {
      sqrt(mean((df$x - df$y.data)^2, na.rm = TRUE))
    })
)
```


```{r, echo=FALSE, fig.height=5,fig.width=9,out.width="100%", fig.align='center', fig.alt='The plot shows the Root Mean Square Error score for prediction horizons from 0 to 25.'}
ggplot(data=rmse) +
  geom_line(aes(k.ahead, rmse), color="steelblue") + 
  geom_point(aes(k.ahead, rmse), color="red") +
  labs(
    title = "Root-Mean Square Errors for Different Prediction Horizons",
    x = "Prediction Steps",
    y = "Root-Mean-Square Errors"
  ) +
  ctsmTMB:::getggplot2theme()
```

## Arguments

---

The `predict` method accepts the following arguments

```{r, eval=FALSE}
model$predict(data,
              pars = NULL,
              method = "ekf",
              ode.solver = "rk4",
              ode.timestep = diff(data$t),
              k.ahead = nrow(data)-1,
              return.k.ahead = 0:min(k.ahead, nrow(data)-1),
              return.variance = c("marginal", "none", "covariance", "correlation"),
              ukf.hyperpars = c(1, 0, 3),
              initial.state = self$getInitialState(),
              estimate.initial.state = private$estimate.initial,
              use.cpp = TRUE,
              silent = FALSE,
              ...)
```

### `pars`

This argument is a vector of parameter values, which are used to generate the predictions. The default behaviour is to use the parameters obtained from the latest call to `estimate` (if any) or alternative to use the initial guesses defined in `setParameter`.

---

### `use.cpp`

Perform predictions in R-based code (`use.cpp=FALSE`) or C++-based code or **C++** (`use.cpp=TRUE`, default).

---

### `method`

See the description in the [estimate vignette](https://phillipbvetter.github.io/ctsmTMB/articles/estimate.html).

---

### `ode.solver`

See the description in the [estimate vignette](https://phillipbvetter.github.io/ctsmTMB/articles/estimate.html).

**Note:** When the argument `use.cpp=TRUE` then the only solvers available are *euler* and *rk4*.

---

### `ode.timestep`

See the description in the [estimate vignette](https://phillipbvetter.github.io/ctsmTMB/articles/estimate.html).

---

### `k.ahead`

This integer argument determines the number of prediction steps desired.

---


### `return.k.ahead`

This vector of integers determines which k-step predictions are returned. The default behaviour is to return all prediction steps (as determined by `k.ahead`).

---

### `return.variance`

The four options are:

1. 'marginal' - return marginal variances only.
2. 'none' - return no dispersion-related quantities.
3. 'covariance' - return full covariance matrix.
4. 'correlation' - return full correlation matrix.

---

### `initial.state`

Sets the initial state $\hat{x}_{0 \mid 0}$ and variance $P_{0 \mid 0}$.

---

### `estimate.initial.state`

A boolean to indicate whether or not estimate the initial state mean value instead of using the one provided via the `model$setInitialState` method or the `initial.state` argument to this method.

The estimation is carried out by root-finding the stationary mean equation equivalent to minimizing

$$
\hat{x}_{0 \mid 0} = \min_{x} \left(f(t_0, x, u_{0}, \theta)\right)^{2}
$$

**Note**: This option is only available when `use.cpp=FALSE` using the pure R implementation.

---

### `silent`

See the description in the [estimate vignette](https://phillipbvetter.github.io/ctsmTMB/articles/estimate.html).

---
