---
title: "simulate: Forecasting the model states and observations."
author: ""
date: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulate: Forecasting the model states and observations.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(ctsmTMB)
```

We demonstate how to use the `simulate` method to obtain k-step ahead predictions / forecasts of a model object, and the observations.

A prediction in this context is considered to be a prior estimate of the state mean and covariance i.e. an estimate obtained without incorporating information from future observations. We denote these

$$
\hat{x}_{i|j} = \mathrm{E}(x_{t_{i}} | x_{t_{j}}, y_{t_{j}}) \\
\hat{P}_{i|j} = \mathrm{V}(x_{t_{i}} | x_{t_{j}}, y_{t_{j}})
$$
where $i \geq j$.

The `predict` method computes the predictions depending on its argument `method` in the following way:

1. The Extended Kalman FIlter (`method=ekf`): The posterior values $\hat{x}_{i|i}$, $\hat{P}_{i|i}$ are obtained from the filter. In each iteration the moment ODEs
$$
\begin{align}
\frac{d\hat{x}_{i|i}}{dt} &= f(\hat{x}_{i|i},\cdots) \\
\frac{d\hat{P}_{i|i}}{dt} &= \left[\frac{df}{dx}(\hat{x}_{i|i},\cdots)\right] \hat{P}_{i|i} + \hat{P}_{i|i} \left[\frac{df}{dx}(\hat{x}_{i|i},\cdots)\right]^{T} + G(\hat{x}_{i|i},\cdots)G^{T}(\hat{x}_{i|i},\cdots)
\end{align}
$$
are solved from the initial time $t_{i}$ until $t_{j}$ where the number of steps is determined by the argument `k.ahead` to `predict`, i.e. `j=i+k.ahead`. The solution at each time point $t_{i+1}, t_{i+2}, \dots, t_{i+k.ahead}$ is the desired predictions. We note that these predictions assume that the underlying probability density of the SDE remains approximately Gaussian. If the prediction horizon is large then this assumption probably doesn't hold, and one should instead use the `simulate` method. The stochastic predictions are generated by simulating the SDE forward in time for a user-defined number of simulations.
 
2. The Laplace Approximation (`method=laplace`): The `predict` method is not yet implemented.

# Simulate from the Ornstein-Uhlenbeck process

---

We use the Ornstein-Uhlenbeck process again.

$$
\mathrm{d}X_{t} = \theta (\mu - X_{t}) \, \mathrm{d}t \, + \sigma_{X} \, \mathrm{d}B_{t} 
$$


$$
Y_{t_{k}} = X_{t_{k}} + e_{t_{k}}, \qquad e_{t_{k}} \sim \mathcal{N}\left(0,\sigma_{Y}^{2}\right)  
$$
We first create data by simulating the process

```{r}
# Simulate data using Euler Maruyama
set.seed(10)
theta=10; mu=1; sigma_x=1; sigma_y=1e-1
# 
dt.sim = 1e-3
t.sim = seq(0,1,by=dt.sim)
dw = rnorm(length(t.sim)-1,sd=sqrt(dt.sim))
x = 3
for(i in 1:(length(t.sim)-1)) {
  x[i+1] = x[i] + theta*(mu-x[i])*dt.sim + sigma_x*dw[i]
}

# Extract observations and add noise
dt.obs = 1e-2
t.obs = seq(0,1,by=dt.obs)
y = x[t.sim %in% t.obs] + sigma_y * rnorm(length(t.obs))

# Create data
.data = data.frame(
  t = t.obs,
  y = y
)
```


# Construct model object

---

We now construct the `ctsmTMB` model object

```{r}
# Create model object
obj = ctsmTMB$new()

# Set name of model (and the created .cpp file)
obj$setModelname("ornstein_uhlenbeck")

# Add system equations
obj$addSystem(
  dx ~ theta * (mu-x) * dt + sigma_x*dw
)

# Add observation equations
obj$addObs(
  y ~ x
)

# Set observation equation variances
obj$setVariance(
  y ~ sigma_y^2
)

# Specify algebraic relations
obj$setAlgebraics(
  theta ~ exp(logtheta),
  sigma_x ~ exp(logsigma_x),
  sigma_y ~ exp(logsigma_y)
)

# Specify parameter initial values and lower/upper bounds in estimation
obj$setParameter(
  logtheta   = log(c(initial = 5,    lower = 0,    upper = 20)),
  mu         = c(    initial = 0,    lower = -10,  upper = 10),
  logsigma_x = log(c(initial = 1e-1, lower = 1e-5, upper = 5)),
  logsigma_y = log(c(initial = 1e-1, lower = 1e-5, upper = 5))
)

# Set initial state mean and covariance
obj$setInitialState(list(x[1], 1e-1*diag(1)))
```

```{r, include=FALSE}
obj$setCppfilesDirectory("../vignettes_sourcefiles")
```

```

# Predict

---

We can use the predict function without having called `estimate` first, in which case the initial value of the parameters provided during `add_parameters` are used.

```{r, results='hide', message=FALSE}
pred = obj$predict(.data)
```

The output of `predict` is a list of two data.frames, one for states and one for observations. The two data.frames share the same five initial columns that contain indices `i` and `j`, associated time-points `t.i` and `t.j`, and a `k.ahead`

They both contain indices, time-points and associated k-step ahead 
, state/observation values and associated standard deviations, and observations provided in the data, and predicted observation values based from the predicted state values.
```{r}
head(pred$states) 
head(pred$observations)
```
The indices `i.` and `j.` in the table refer to expectation and variance of the state vector
$$
\mathrm{E}(x_{t_{i}} | x_{t_{j}}, y_{t_{j}}) \\
\mathrm{V}(x_{t_{i}} | x_{t_{j}}, y_{t_{j}})
$$
where the moments are given at indices $i$ (time $t_{i}$) conditioned on the information available up until indices $j$ (time $t_{j}$). In this sense the state/covariance predictions with $i=j$ are posterior estimates, and all others are prior estimates.

By default only the standard deviations of the states are returned. The entire covariance matrix elements can be returned via
```{r, eval=FALSE}
obj$predict(.data, return.covariance = TRUE)
```


# Set number of prediction steps

---

The number of time steps ahead that predictions are desired can be changed with the `k.ahead` argument. The default behaviour is to save all prediction steps from 1 to `k.ahead` but if only a selected few are desired, then the `return.k.ahead` argument can be used to indicate which should be kept in the output.
```{r, results='hide', message=FALSE}
pred1 = obj$predict(.data, k.ahead=2)
pred2 = obj$predict(.data, k.ahead=10, return.k.ahead=c(2,5,8))
```
```{r}
head(pred1$states)
```
```{r}
head(pred2$states)
```



# Set the model parameters used

---

The default of `predict` is to use the initial parameters supplied during `setParameter`, unless the `estimate` function has been succesfully run, then `predict` will use the parameter values at the found minimizer. You can provide other parameters using the `pars` argument of the method i.e.
```{r, eval=FALSE}
some.other.random.pars = rnorm(length(fit$par.fixed))
pred = obj$predict(.data, pars=some.other.random.pars)
```

# Set the initial state and covariance

---

The default behaviour of `predict` is to use the initial state and covariance suppled when calling the `set_initial_state` method. This can be important to change if one wishes to correctly predict on the first few observations in a provided prediction data series. You can supply the `x0` and `p0` arguments with updated state and covariance estimates
```{r, eval=FALSE}
new.initial.x0 = rnorm(1)
new.initial.p0 = rnorm(1)*diag(1)
pred = obj$predict(data=.data,
                   initial.state = list(
                     new.state.value,
                     new.covariance.value
                   )
)
```

# The solver options

---

You can choose which solver options to use via the arguments `ode.solver` and `ode.timestep`. 

The `ode.solver` decides which ode solver algorithm to used when solving the mean and variance ODEs of the SDE. 

The `ode.timestep` determines the time step-size using when solving the ODEs, with default value being the minimum observed time difference in the provided data time vector (`min(diff(.data$t))`). The time step-size can't be smaller than this. If the chosen time step-size $\Delta_{step}$ does not "almost" divide the observation time differences in the data $\Delta_{obs}$ i.e. if 
$$
\frac{\Delta_{obs}}{\Delta_{step}} - \bigg\lfloor \frac{\Delta_{obs}}{\Delta_{step}} \bigg\rfloor \geq 0.02
$$
then the time step-size is reduced such that it does divide evenly by setting

$$
\frac{\Delta_{obs}}{\Delta_{step}} = \bigg\lceil \frac{\Delta_{obs}}{\Delta_{step}} \bigg\rceil
$$
In other words: if there it takes $3.561$ time-steps to get from $t_{i}$ to $t_{i+1}$ then we take $4$ time-steps instead, and calculate a reduced time-step such that this is true. If there were only $3.01$ time-steps then we retain the original time-step, ignoring the small temporal discrepancy. 

We provide these arguments via
```{r, eval=FALSE}

obj$predict(.data,
            ode.timestep = min(diff(.data$t))
)
```

# Use-case

---

We could use our model predictions against observation to compute a model performance score e.g. RMSE (root-mean square error). We first estimate the model parameters, such that they are automatically used when calling `predict`. We then predict up to 10-steps ahead, and return only these values.

```{r, results='hide', message=FALSE}
fit = obj$estimate(.data)
pred = obj$predict(.data, k.ahead=10)
pred.states = pred$states
pred.obs = pred$observations
```

```{r, message=FALSE, results='hide'}
# ggplot2 theme
library(ggplot2)
mytheme =
  theme_minimal() + 
  theme(
    text             = element_text("Avenir Next Condensed",size=15),
    legend.text      = element_text(size=15),
    axis.text        = element_text(size=15),
    strip.text       = element_text(face="bold",size=15),
    legend.box       = "vertical",
    legend.position  = "top",
    plot.title       = element_text(hjust=0.5)
  )
```

Let's plot the 10-step predictions against the observations.
```{r}
pred10 = pred.states[pred.states$k.ahead==10,]
```


```{r}
data = .data
ggplot() +
  geom_line(aes(x=pred10$t.j,y=pred10$x,color="10-Step Predictions")) +
    geom_ribbon(aes(x=pred10$t.j,ymin=pred10$x-2*sqrt(pred10$var.x),ymax=pred10$x+2*sqrt(pred10$var.x)),fill="grey",alpha=0.5) +
  geom_point(aes(x=data$t,data$y,color="Observations")) +
  labs(color="",x="Time",y="") +
  # coord_cartesian(xlim=c(0,0.1)) +
  mytheme
```
We notice that because these are 10-step predictions the state uncertainty (95\%) is much larger than for the filtered posterior state estimates from `fit` i.e.

```{r}
t         = fit$states$mean$posterior$t
xpost     = fit$states$mean$posterior$x
xpost_sd  = fit$states$sd$posterior$x

ggplot() +
  geom_line(aes(x=t,y=xpost,color="0-Step Predictions (Posterior State Estimates)"),lwd=1) +
  geom_ribbon(aes(x=t,ymin=xpost-2*xpost_sd,ymax=xpost+2*xpost_sd),fill="grey",alpha=0.5) +
  geom_point(aes(x=data$t,data$y,color="Observations")) +
  labs(x = "Time", y = "", color="") +
  mytheme
```
We can calculate the RMSE prediction score for each prediction step as follows:
```{r}
rmse = c()
k.ahead = 1:10
for(i in k.ahead){
  xy = data.frame(
    x = pred.states[pred.states$k.ahead==i,"x"],
    y = pred.obs[pred.obs$k.ahead==i,"y.data"]
  )
  rmse[i] = sqrt(mean((xy[["x"]] - xy[["y"]])^2))
}

ggplot() +
  geom_line(aes(k.ahead, rmse), color="steelblue") + 
  geom_point(aes(k.ahead, rmse), color="red") +
  labs(
    title = "Root-Mean Square Errors for Different Prediction Horizons",
    x = "Prediction Steps",
    y = "Root-Mean-Square Errors"
    ) +
  mytheme
```
